### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving articial absorption into an appendix and the reduced domain approximation into the spn chapter

For Jacobi-based preconditioners, convergence with MCSA is difficult
if not impossible for ill-conditioned systems with this behavior
demonstrated using a simple neutron diffusion problem. In order to
effectively solve the $SP_N$ equations for the fuel assembly problem,
a suite of preconditioners along with a relaxation scheme is developed
and studied within the context of MCSA. Using these preconditioners,
several additional issues are observed and alleviated to a certain
extent by applying the reduced domain approximation. MCSA solutions
for the fuel assembly problem are then verified by comparing the
solutions against production Krylov solvers for problems with varying
numbers of energy groups. Finally performance is analyzed through
comparison with those same Krylov solvers in terms of both iterative
performance and CPU timing using the same set of problems utilized for
the verification.

@@

\subsection{Block Jacobi Preconditioning}
\label{subsec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their diffusion-like form. In this section, we will
the structure of the transport operator formed by the $SP_N$ equations
while parametrically varying the $SP_N$ order, the $P_N$ order, the
number of energy groups, and upscattering/downscattering between
energy groups. For each basic problem, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively.

@@

We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
$N_g\times(N+1)/2$.

@@

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
A single energy group problem was first computed with $SP_1$
discretization, effectively giving the one-speed neutron diffusion
system for the fuel assembly resulting in 20,088 degrees of freedom in
the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. To
investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff as no
artificial absorption was used for these preliminary
calculations. Based on this, the correction being generated by the
Neumann-Ulam solve is the component of MCSA causing a divergent
solution which we study next.

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi preconditioned
calculations used a one-speed $SP_1$ discretization, effectively
giving a diffusion system. To study the breakdown of MCSA at iteration
matrix spectral radii near one, we will use the simpler homogeneous
2-dimensional one-speed neutron diffusion system presented in
Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
this system, we can vary the cross sections while maintaining a fixed
grid in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

@@

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing as a function of spectral radius. Near a spectral radius of
0.97, the number of histories required to converge MCSA in less than
100 iterations takes a dramatic rise that exhibits neither exponential
nor power law behavior. As the spectral radius approaches 1, the
number of histories required becomes significant and effectively
impractical to compute. Even with this simple diffusion problem, the
behavior is consistent with that observed for the fuel assembly
problem with $SP_1$ discretization. In that case, we estimated a
spectral radius of $\approx 0.998$, larger than any of the spectral
radii that could be computed within even 90 minutes of compute time
for this simple two dimensional problem. For that problem, even
1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
convergence. Even if single solve times of an hour can be tolerated,
dozens of solves are typically required to compute the multiplication
factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
difficult problems like the fuel assembly, making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}

In addition to the significantly larger number of histories required
to achieve convergence for ill-conditioned problems another penalty is
paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. At
iteration matrix spectral radii this large, using significant numbers
of stochastic histories in the Monte Carlo solve is not enough to
reduce the stochastic uncertainty in the correction, causing the MCSA
sequence to diverge instead of accelerating convergence. We therefore
require a more expansive set of preconditioning techniques to move the
eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
is more applicable and in which performance is improved.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

@@

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
ILUT preconditioning is chosen first due to its applicability to
problems that have a dominating elliptic component such as the fuel
assembly problem or other problems with a high scattering to
absorption ratio. As with MCSA, the algebraic formulation of ILUT
requires the physics operator to be fully formed and is therefore
easily incorporated into our solution sequence.

@@

\subsection{Applying the Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
Although ILUT and SPAINV preconditioning permitted MCSA solutions for
the fuel assembly problem, a primary concern is the number of non-zero
states in each row of the system generated by the explicit
preconditioning strategy. In many cases, orders of magnitude more
matrix elements were generated resulting in poor scalability for
domain decomposed algorithms and overall performance issues for Monte
Carlo. As outlined in \S~\ref{subsec:reduced_domain_approximation},
the reduced domain approximation may be used as a mechanism to
potentially alleviate this problem by filtering elements of the
composite matrix in each row that fall below a certain threshold value
or by maintaining the largest $N$ elements in each row where $N$ is a
designated fill level. In this section, we will apply the reduced
domain approximation to both ILUT and SPAINV preconditioned system to
reduce the density of the composite linear operator to more manageable
levels and study the resulting iterative performance.

For each preconditioner, the parameters that achieved the best quality
metric results from the previous analysis were used. These correspond
to ILUT parameters of a fill level of 5.0 and a drop tolerance of
\sn{1}{-5} and SPAINV parameters of a sparsity level of 4 and a
threshold of 0.1. The reduced domain threshold was set to \sn{1}{-10}
in order to eliminate any exceedingly small values from the matrix
(often this is simply removing non-zero elements within the floating
point tolerance of zero). The reduced domain fill level was then
varied, starting with the largest non-zero entries per row value
observed for each of the preconditioning types in order to assess its
effects relative to the case where no reduced domain approximation was
applied.

@@

Although applying the reduced domain approximation results in a
successful recovery of sparsity for the Monte Carlo problem while
maintaining good convergence properties, there is still an issue of
forming the composite operator before applying the approximation and
potentially generating the transpose of this operator in the case of
the adjoint Monte Carlo method. Because of this, memory and scaling
issues may still be observed when building the probability and weight
matrices for the Monte Carlo problem. In addition, the expensive
extraction of the inverse of the preconditioning operators for the
explicit scheme creates a significant cost in overall
performance. Future work in this area should be considerate of these
important components of the preconditioned MCSA algorithm.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Verification\ }
\label{sec:spn_mcsa_verification}
Through the numerical studies presented in this chapter, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. In addition,
several steps were taken to mitigate the dense composite linear
operators that arise when performing explicit MCSA preconditioning
necessary for achieve convergence. Using these results, we now compare
MCSA to conventional Krylov solvers that would be typically used to
solve the $SP_N$ form of the transport equation as presented here.

MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
groups using $SP_1$ discretization will computed using both BiCGStab
and GMRES implementations from the Trilinos scientific computing
library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
number of eigenvalue iterations required to converge the problem, and
the eigenvalue itself will be compared as a means of verification. For
the flux values, we will look at the 2-norm and infinity norm of the
relative difference vector between the reference case and the
comparison case in each energy group:
\begin{equation}
  \text{Relative Group Flux Difference} = \frac{\ve{\Phi}_{ref}^g -
    \ve{\Phi}^g}{\ve{\Phi}_{ref}^g}\:,
  \label{eq:relative_difference_vector}
\end{equation}
where the addition and subtraction operations are performed
element-wise to construct the vector. For this work, the BiCGStab
results will be taken as the reference case.

All solvers will be preconditioned with ILUT using a drop tolerance of
\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
placed on the size of the subspace and therefore no restarts
occurred. When the collision estimator was used with MCSA, \sn{2}{4}
stochastic histories were used to compute the correction for every
energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
for the 2 and 4 group calculations respectively) corresponding to
approximately 1 stochastic history per DOF. When the expected value
estimator was used, \sn{5}{2} stochastic histories used for each
energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
4 group calculations) giving approximately 1 stochastic history for
every 5 DOFs. The reduced domain approximation was also applied in
conjunction with the ILUT preconditioning using a fill level of 100
and a threshold value of \sn{1}{-10} to reduce the density of states
in the Monte Carlo problem. For relaxation parameters, all MCSA
computations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator. Additionally, both the
Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa}) and the
minimal residual-based MCSA iteration given by
Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary performance
analysis for this set of MCSA variations is presented in
Appendix~\ref{chap:spn_estimator_comparison}.

Table~\ref{tab:spn_solver_defs} gives definitions for the
solvers used to generate the results in the remainder of this
section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
group 0).

@@

Table~\ref{tab:serial_differences_g1},
Table~\ref{tab:serial_differences_g2}, and
Table~\ref{tab:serial_differences_g4} give the infinity norms and
2-norms of the flux relative difference vectors given by
Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
energy groups. In general, we see good agreement in the 2-norm between
the MCSA variants and the GMRES calculation, meaning that they have
approximately the same variation from the BiCGStab calculation. We do
note larger infinity norm values of this vector for some of the MCSA
variants relative to the GMRES results (i.e. group 1 in the 2-group
calculation), however, this large maximum value is apparently a local
phenomenon with good agreement in the 2-norm for the
calculations. Based on the good agreement between MCSA and GMRES for
the relative difference of the group fluxes and the eigenvalue results
presented here, the MCSA implementation should be deemed correct for
the fuel assembly problem for serial calculations with multiple energy
groups.

>>>>

For Jacobi-based preconditioners, convergence with MCSA is difficult
if not impossible for ill-conditioned systems with this behavior
demonstrated using a simple neutron diffusion problem. In order to
effectively solve the $SP_N$ equations for the fuel assembly problem,
a suite of preconditioners along with a relaxation scheme is developed
and studied within the context of MCSA. Using these preconditioners,
several additional issues are observed and alleviated to a certain
extent by developing and applying the reduced domain
approximation. MCSA solutions for the fuel assembly problem are then
verified by comparing the solutions against production Krylov solvers
for problems with varying numbers of energy groups. Finally
performance is analyzed through comparison with those same Krylov
solvers in terms of both iterative performance and CPU timing using
the same set of problems utilized for the verification.

@@

\subsection{Block Jacobi Preconditioning}
\label{subsec:spn_spectral_analysis}
To apply MCSA to the fuel assembly problem, we require some form of
preconditioning to achieve an iteration matrix spectral radius of less
than unity. Previous work in applying MCSA to the thermal radiation
diffusion equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their diffusion-like form. We first investigate the
structure of the transport operator formed by the $SP_N$ equations
while parametrically varying the $SP_N$ order, the $P_N$ order, the
number of energy groups, and upscattering/downscattering between
energy groups with the values given in
Table~\ref{tab:spn_fixed_parameters} fixed for each spatial location
and energy group.

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. We note a few key features of the sparsity
plots. The first is that for multigroup problems without full
upscatter and downscatter (i.e. Figure~\ref{fig:group10ds}), the
resulting matrix is asymmetric and therefore a linear solver that can
handle asymmetric linear systems is required. Nearly all problems of
interest will not have full upscattering or downscattering. Second we
note the largely diagonal character of these systems, although the
blocks from Eq~(\ref{eq:A_block_matrix}) are readily apparent with a
block size of $N_g\times(N+1)/2$.

@@

@@

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
A single energy group problem was first computed with
$SP_1$ discretization, effectively giving the one-speed neutron
diffusion system for the fuel assembly resulting in 20,088 degrees of
freedom in the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives
the residual infinity norm as a function of iteration for the MCSA
linear solve in the first eigenvalue iteration using 25,000 stochastic
histories at every iteration for the adjoint Neumann-Ulam solve.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}

Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. To
investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}

Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff. Based on
this, the correction being generated by the Neumann-Ulam solve is the
component of MCSA causing a divergent solution which we study next.

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
To study the breakdown of MCSA at iteration matrix spectral radii near
one, we will use the simpler homogeneous 2-dimensional one-speed
neutron diffusion system presented in
Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
this system, we can vary the cross sections while maintaining a fixed
grid in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

@@

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing as a function of spectral radius. Near a spectral radius of
0.97, the number of histories required to converge MCSA in less than
100 iterations takes a dramatic rise that exhibits neither exponential
nor power law behavior. As the spectral radius approaches 1, the
number of histories required becomes significant and effectively
impractical to compute. Even with this simple diffusion problem, the
behavior is consistent with that observed for the fuel assembly
problem with $SP_1$ discretization. In that case, we estimated a
spectral radius of $\approx 0.998$, larger than any of the spectral
radii that could be computed within even 90 minutes of compute time
for this simple two dimensional problem. For that problem, even
1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
convergence. Even if single solve times of an hour can be tolerated,
dozens of solves are typically required to compute the multiplication
factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
difficult problems like the fuel assembly, making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}

In addition to the significantly larger number of histories required
to achieve convergence for ill-conditioned problems another penalty is
paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. At
iteration matrix spectral radii this large, using significant numbers
of stochastic histories in the Monte Carlo solve is not enough to
reduce the stochastic uncertainty in the correction, causing the MCSA
sequence to diverge instead of accelerating convergence. We therefore
require a more expansive set of preconditioning techniques to move the
eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
is more applicable and in which performance is improved.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

@@

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
ILUT preconditioning is chosen first due to its applicability to
problems that have a dominating elliptic component such as the fuel
assembly problem or other problems with a large scattering to
absorption ratio. The algebraic formulation of ILUT requires the
physics operator to be fully formed, also a requirement for MCSA, and
is therefore easily incorporated into our solution sequence.

@@

\subsection{The Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
Although ILUT and SPAINV preconditioning permitted MCSA solutions for
the fuel assembly problem, a primary concern is the number of non-zero
states in each row of the system generated by the explicit
preconditioning strategy. In many cases, orders of magnitude more
matrix elements were generated resulting in poor scalability for
domain decomposed algorithms and overall performance issues for Monte
Carlo. To alleviate this, we propose modifying the domain sampled by
the Monte Carlo method to generate the correction within MCSA by
removing terms in the iteration matrix through some predetermined
criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
stochastic realization of multiple matrix-vector multiplies, the
criteria we choose should be based on maintaining those elements that
will make the largest contributions to the multiplication and
therefore recover as much of the original product as possible.

For each row in the preconditioned system, we can then apply the
reduced domain approximation to eliminate those elements that either
fall below a specified tolerance level, keep the $N$ largest elements
in each row, or both giving:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
    \label{eq:rda_correction}\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:rda_mcsa}
\end{subequations}
where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
reduced domain approximation has been applied to the iteration matrix
in the Monte Carlo method. If enough of the terms in the original
iteration matrix have been maintained, the character of the original
iteration matrix should be retained and continue to accelerate the
iterative scheme.

We next apply the reduced domain approximation to both ILUT and SPAINV
preconditioned system to reduce the density of the composite linear
operator to more manageable levels and study the resulting iterative
performance. For each preconditioner, the parameters that achieved
the best quality metric results from the previous analysis were
used. These correspond to ILUT parameters of a fill level of 5.0 and a
drop tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level
of 4 and a threshold of 0.1. The reduced domain threshold was set to
\sn{1}{-10} in order to eliminate any exceedingly small values from
the matrix (often this is simply removing non-zero elements within the
floating point tolerance of zero). The reduced domain fill level was
then varied, starting with the largest non-zero entries per row value
observed for each of the preconditioning types in order to assess its
effects relative to the case where no reduced domain approximation was
applied.

@@

Although applying the reduced domain approximation results in a
successful recovery of sparsity for the Monte Carlo problem while
maintaining good convergence properties with ILUT, there is still an
issue of forming the composite operator before applying the
approximation and potentially generating the transpose of this
operator in the case of the adjoint Monte Carlo method. Because of
this, memory and scaling issues may still be observed when building
the probability and weight matrices for the Monte Carlo problem. In
addition, the expensive extraction of the inverse of the
preconditioning operators for the explicit scheme creates a
significant cost in overall performance. Future work in this area
should be considerate of these important components of the
preconditioned MCSA algorithm.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Verification\ }
\label{sec:spn_mcsa_verification}
Through the numerical studies presented in this chapter, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. In addition,
several steps were taken to mitigate the dense composite linear
operators that arise when performing explicit MCSA preconditioning
necessary for achieve convergence. Using these results, we now compare
MCSA to conventional Krylov solvers that would be typically used to
solve the $SP_N$ form of the transport equation as a means of
verification.

MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
groups using $SP_1$ discretization will computed using both BiCGStab
and GMRES implementations from the Trilinos scientific computing
library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
number of eigenvalue iterations required to converge the problem, and
the eigenvalue itself will each be evaluated. For the flux values, we
will look at the 2-norm and infinity norm of the relative difference
vector between the reference case and the comparison case in each
energy group:
\begin{equation}
  \text{Relative Group Flux Difference} = \frac{|\ve{\Phi}_{ref}^g -
    \ve{\Phi}^g|}{|\ve{\Phi}_{ref}^g|}\:,
  \label{eq:relative_difference_vector}
\end{equation}
where the absolute value, subtraction, and division operations are
performed element-wise to construct the vector. For this work, the
BiCGStab results will be taken as the reference case.

All solvers will be preconditioned with ILUT using a drop tolerance of
\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
placed on the size of the subspace and therefore no restarts
occurred. When the collision estimator was used with MCSA, \sn{2}{4}
stochastic histories were used to compute the correction for every
energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
for the 2 and 4 group calculations respectively) corresponding to
approximately 1 stochastic history per DOF. When the expected value
estimator was used, \sn{5}{2} stochastic histories used for each
energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
4 group calculations) giving approximately 1 stochastic history for
every 5 DOFs. The reduced domain approximation was also applied in
conjunction with the ILUT preconditioning using a fill level of 100
and a threshold value of \sn{1}{-10} to reduce the density of states
in the Monte Carlo problem. For relaxation parameters, all MCSA
computations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator. When the collision estimator
was used, an artificial absorption value of 0.2 as outlined in
Appendix~\ref{chap:artificial_absorption} was also
applied. Additionally, both the Richardson-based MCSA iteration given
by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
given by Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary
performance analysis for this set of MCSA variations is presented in
Appendix~\ref{chap:spn_estimator_comparison}. Table~\ref{tab:spn_solver_defs}
gives definitions for the solvers used to generate the results in the
remainder of this section and Table~\ref{tab:spn_group_structure}
gives the group structures with lower bounds of each group in eV.

@@

Table~\ref{tab:serial_differences_g1},
Table~\ref{tab:serial_differences_g2}, and
Table~\ref{tab:serial_differences_g4} give the infinity norms and
2-norms of the flux relative difference vectors given by
Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
energy groups. In general, we see good agreement in the 2-norm between
the MCSA variants and the GMRES calculations, meaning that they have
approximately the same variation from the BiCGStab calculations. We do
note larger infinity norm values of this vector for some of the MCSA
variants relative to the GMRES results (i.e. group 1 in the 2-group
calculation), however, this large maximum value is apparently a local
phenomenon with good agreement in the 2-norm for the
calculations. Based on the good agreement between MCSA and GMRES for
the relative difference of the group fluxes and the eigenvalue results
presented here, the MCSA implementation should be deemed correct for
the fuel assembly problem for serial calculations with multiple energy
groups.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lccc}\hline\hline
      \multicolumn{1}{c}{\textbf{Solver}} & 
      \multicolumn{1}{c}{\textbf{Group}} &
      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
      \hline
      GMRES-ILUT & 0 & \sn{4.562}{-6} & \sn{6.746}{-5} \\
      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} & \sn{2.541}{-4} \\
      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} & \sn{4.672}{-4} \\
      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} & \sn{2.740}{-4} \\
      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} & \sn{2.554}{-4} \\
      Richardson-ILUT & 0 & \sn{1.531}{-5} & \sn{2.771}{-4} \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Serial scalar flux verification results for the
      1-group fuel assembly problem.} \textit{The infinity norm and
      2-norm of the relative flux vector in each group is given with
      BiCGStab used as the reference.}}
  \label{tab:serial_differences_g1}
\end{table}
